Immutable Source-Mapped Abstract Syntax Tree:\ \\ A Design Pattern for Refactoring Engine APIs

### 2014-05-01 paper.tex
Camera-ready

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree: \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

\begin{abstract}
Many automated refactoring tools provide an extensibility mechanisms that
allows third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  This pattern is used in the Eclipse Java
Development Tools' refactoring API as well as the Microsoft ``Roslyn'' CTP,
among other uses.
\end{abstract}

@@

\terms{Languages}
\keywords{abstract syntax trees, application programming interfaces, APIs,
ASTs, immutability, refactoring}

@@

\begin{bottomstuff}
This material is based upon work supported by the National Science Foundation
under Grant No.~CCF-1217271.

Permission to make digital or hard copies of all or part of this work for
personal or classroom use is granted without fee provided that copies are not
made or distributed for profit or commercial advantage and that copies bear
this notice and the full citation on the first page. To copy otherwise, to
republish, to post on servers or to redistribute to lists, requires prior
specific permission. A preliminary version of this paper was presented in a
writers' workshop at the 20th Conference on Pattern Languages of Programs
(PLoP).
\end{bottomstuff}

@@

\begin{wrapfigure}{r}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.62]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST produced by the Eclipse Java Development Tools
(JDT).}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
program:
\vspace*{1em}

@@

\vspace*{1em}
Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

@@

Deeper discussions on ASTs can be found in many introductory textbooks on
compiler construction (e.g., \cite{aho06compilers} and
\cite{torczon11engineering}).  Such books also discuss how parsers are
constructed, including how they can be modified to construct ASTs.

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  After all, immutable
objects tend to have simpler interfaces, and they are thread-safe.  Both
authors also emphasize that APIs should be easy to learn (even for programmers
who do not read all of the API documentation) and difficult for clients to
misuse.  Given these guidelines, and the list of liabilities for mutable ASTs
given above, it seems preferable from the perspective of someone attempting to
publish an API to avoid mutable ASTs.  But this raises a different question: If
ASTs are immutable, what API should be provided for clients to manipulate
source code?

@@

In their discussions of API design, Bloch submits that it should be ``easy to
read and maintain code that uses [the API]''~\cite{bloch}, and Tulach notes the
advantages of APIs where the clients ``don't describe step-by-step what they
want your API to do.  Rather, they `declare' what they want to have happen and
then rely on your API to do it\dots''~[p.~225]\cite{tulach08practical}.

@@

Eclipse JDT provides a class (\textit{ASTRewrite}\footnote{\url{http://help.eclipse.org/kepler/topic/org.eclipse.jdt.doc.isv/reference/api/org/eclipse/jdt/core/dom/rewrite/ASTRewrite.html}}) that allows the programmer
to specify modifications in terms of Java AST nodes, and then translates those
changes into an LTK \textit{TextEdit} object.  Eclipse CDT's refactoring engine
is based on JDT's and follows a similar design; it offers its own version of
textit{ASTRewrite}~\cite{cdt-refactoring}.

@@

Microsoft's ``Roslyn'' CTP takes a functional approach, where imperative
methods on AST nodes do not modify the AST but rather return a new
AST~\cite{vogel12roslyn}.  Since a na\"{i}ve implementation could make this
prohibitively expensive, the Roslyn team devised an implementation that
they call \textit{red-green trees} (named after the colors of the
whiteboard markers that were used during the design
meeting)~\cite{lippert12persistence}.  Essentially, the AST exposed via the API
(the ``red tree'') is a fa\,{c}ade for a different, internal tree (the ``green
tree''); the implementation attempts to reuse existing AST nodes whenever
possible (which is possible since all nodes are immutable).

@@

\textit{The author would like to thank Peter Sommerlad for the invaluable
feedback and suggestions he provided as the shepherd for this paper.}

>>>>

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

\begin{abstract}
Many automated refactoring tools provide an extensibility mechanisms that
allows third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  In particular, these trees tend to have an
immutable structure, and they maintain very detailed information about the
source code, including exact character positions.  This pattern is used in the
Eclipse Java Development Tools' refactoring API as well as the Microsoft
``Roslyn'' CTP, among other uses.
\end{abstract}

@@

\terms{Languages}
\keywords{abstract syntax trees, APIs, ASTs, immutability, refactoring}

@@

\begin{bottomstuff}
Permission to make digital or hard copies of all or part of this work for
personal or classroom use is granted without fee provided that copies are not
made or distributed for profit or commercial advantage and that copies bear
this notice and the full citation on the first page. To copy otherwise, to
republish, to post on servers or to redistribute to lists, requires prior
specific permission. A preliminary version of this paper was presented in a
writers' workshop at the 20th Conference on Pattern Languages of Programs
(PLoP).  PLoP'13, October 23--26, Monticello, Illinois, USA. Copyright 2013 is
held by the author(s). HILLSIDE 978-1-941652-00-8
\end{bottomstuff}

@@

\begin{wrapfigure}{r}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.62]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST produced by the Eclipse Java Development Tools
(JDT).\vspace*{-4em}}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
program:
\vspace*{.5em}

@@

\vspace*{.5em}
Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

@@

Deeper discussions on ASTs can be found in many introductory textbooks on
compiler construction, such as \cite{aho06compilers} and
\cite{torczon11engineering}.  Such books also discuss how parsers are
constructed, including how they can be modified to construct ASTs.

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  Immutable objects tend
to have simpler interfaces, and they are thread-safe.  Both authors also
emphasize that APIs should be easy to learn (even for programmers who do not
read all of the API documentation) and difficult for clients to misuse.  Given
these guidelines, and the list of liabilities for mutable ASTs given above, it
seems preferable from the perspective of someone attempting to publish an API
to avoid mutable ASTs.  But this raises a different question: If ASTs are
immutable, what API should be provided for clients to manipulate source code?

@@

In their discussions of API design, Bloch submits that it should be ``easy to
read and maintain code that uses [the API]''~\cite{bloch}, and Tulach notes the
advantages of APIs where the clients ``don't describe step-by-step what they
want your API to do.  Rather, they `declare' what they want to have happen and
then rely on your API to do it\dots''~\cite[p.~225]{tulach08practical}.

@@

Eclipse JDT provides a class (\textit{ASTRewrite}\footnote{\url{http://help.eclipse.org/kepler/topic/org.eclipse.jdt.doc.isv/reference/api/org/eclipse/jdt/core/dom/rewrite/ASTRewrite.html}}) that allows the programmer
to specify modifications in terms of Java AST nodes, and then translates those
changes into an LTK \textit{TextEdit} object.  Eclipse CDT's refactoring engine
is based on JDT's and follows a similar design; it offers its own version of
\textit{ASTRewrite}~\cite{cdt-refactoring}.

@@

Microsoft's ``Roslyn'' CTP takes a functional approach, where imperative
methods on AST nodes do not modify the AST but rather return a new
AST~\cite{vogel12roslyn}.  Since a na\"{i}ve implementation could make this
prohibitively expensive, the Roslyn team devised an implementation that
they call \textit{red-green trees} (named after the colors of the
whiteboard markers that were used during the design
meeting)~\cite{lippert12persistence}.  Essentially, the AST exposed via the API
(the ``red tree'') is a fa\c{c}ade for a different, internal tree (the ``green
tree''); the implementation attempts to reuse existing AST nodes whenever
possible (which is possible since all nodes are immutable).

@@

\section*{Acknowledgments}
This material is based upon work supported by the National Science Foundation
under Grant No.~CCF-1217271.  The author would like to thank Peter Sommerlad,
who shepherded this pattern, as well as the members of Workshop~000001 at
PLoP~2013: Filipe Correia, Dibyendu Goswami, Joshua Kerievsky, Juan Reza, and
Russ Rubis.  Their feedback was immensely helpful.
